---
title: "Data Wrangling"
subtitle: "a module of transforming data"
output: 
   learnr::tutorial:
     theme: "lumen" 
runtime: shiny_prerendered
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)

(student_data <- tibble(
  Student = c("Abbie", "Billy", "Clara", "Danielle", "Edgar"),
  Math_Score = c(85, 92, 78, 95, 88),
  Literature_Score = c(88, 90, 75, 92, 86)
))
```

## A Brief History of R

### The Birth of R:

R's story is intimately tied to the history of the S programming
language. In the early 1970s, John Chambers and his colleagues at Bell
Laboratories developed S, a proprietary statistical programming
language. S revolutionized data analysis by introducing a structured and
powerful language for statistical computing. While S was groundbreaking,
it was a proprietary software, limiting its accessibility and
affordability. This led to the desire for an open-source alternative,
which ultimately gave birth to R. Ross Ihaka and Robert Gentleman's work
in the early 1990s, which resulted in R, was heavily influenced by S.
Their goal was to develop a free, accessible alternative to the
expensive statistical software packages available at the time.

### Open-Source Roots:

To accomplish the goal of free accessible software, R was developed
under the GNU General Public License (GPL), ensuring it would remain
open and accessible to everyone. R's open-source nature meant that
anyone could contribute to its development and use it for various
statistical and data analysis tasks. This openness quickly attracted a
community of statisticians, data scientists, and researchers who saw its
potential. Their collaborative approach led to rapid improvements, bug
fixes, and the creation of numerous packages that extended R's
capabilities.

### Legacy of S in R:

Despite the transition to R, S's legacy lives on within R. Many
concepts, functions, and syntax from S were integrated into R, making it
familiar to those who had experience with S. This ensured a relatively
smooth transition for statisticians and data analysts. As R gained
popularity and traction within the statistical and data science
communities, it had a notable impact on S. The success of R contributed
to the decline of S as a dominant statistical platform. It became clear
that open-source and community-driven development was the way forward in
the field of statistical computing.

### Evolution and Packages:

As R gained popularity, an ecosystem of user-contributed packages began
to flourish. These packages added specific functions and capabilities to
R, making it even more versatile. The Comprehensive R Archive Network
(CRAN) became the central repository for these packages, ensuring easy
access for users.

### R's Contribution to Data Science:

In the early 2000s, as the field of data science started to emerge, R
gained prominence due to its robust statistical capabilities and data
visualization tools. It became an essential tool for data analysts and
scientists for tasks like data exploration, statistical modeling, and
visualization.

### R's Impact on Academia and Industry:

R's open-source and community-driven nature led to its widespread
adoption in both academia and industry. It became a standard tool for
researchers and data professionals in fields such as economics, biology,
social sciences, finance, and more.

### Integration with Big Data and Machine Learning:

As the data landscape evolved, R adapted. It integrated with big data
technologies like Hadoop and Spark, allowing users to analyze massive
datasets. R also developed machine learning libraries such as caret and
xgboost, further solidifying its role in data science.

### R Today:

Today, R remains a powerful and popular tool for data analysis and
statistics. It has a large and active user community, a vast library of
packages, and is widely taught in college courses and data science
programs.

### Why Learn R:

As college students, learning R can provide you with a solid foundation
in statistical analysis and data manipulation. It's a valuable skill for
future careers in data science, analytics, research, and various other
fields.

In summary, R started as a grassroots effort to democratize statistical
analysis, evolved into a versatile data analysis tool, and continues to
play a pivotal role in data science today. Embracing R as you embark on
your journey into data analysis and statistics can open doors to a world
of analytical possibilities.

## Data Wrangling

Data wrangling is a crucial step in the data analysis process. It
involves transforming, cleaning, and reshaping raw data into a format
that is more suitable for analysis and visualization. Data wrangling
ensures that your data is in the right shape and structure to extract
meaningful insights effectively.

In this tutorial, we will explore data wrangling using Tidyverse, a
collection of R packages designed for data science that includes popular
packages like `dplyr`, `tidyr`, and `ggplot2`. Tidyverse provides a
powerful and consistent set of tools for data manipulation and
visualization, making it a favorite among data analysts and scientists.

The data wrangling process spans several steps of the data science
process, including consists of the following steps:

-   **Import Data:** Importing data from various sources, such as files
    (CSV, Excel), databases, APIs, or web scraping.
-   **Data Cleaning:** Addressing missing or erroneous data, handling
    duplicates, and converting data types to ensure consistency.
-   **Data Transformation:** Reorganizing and restructuring data,
    including tasks like filtering rows, selecting columns, creating new
    variables, and reshaping data when needed.

Throughout this tutorial, we will dive into each of these steps and
demonstrate how Tidyverse's tools can simplify and streamline the data
wrangling process, ultimately helping you derive valuable insights from
your data.

Are you ready to get started with data wrangling? Let's begin!

## The Flavors of R

Welcome to this tutorial on base R and data frames! R has evolved and
improved rapidly over the years. R functions and coding from the early
days (90s) is known as base R. Even though we will be embracing the
tidyverse philosophy of R, bits of base R code shows up regularly in
examples, videos, and code generated by AI. Here is a brief introduction
to base R and how it compares to tidyverse so you know it when you see
it.

R offers several flavors or dialects that cater to different
preferences. The main flavors include "base R," the "Tidyverse," and
"data.table." Base R is the default flavor when you first open R and has
a long-standing history. In contrast, the Tidyverse and data.table are
extensions accessed through packages. Each flavor has its strengths and
applications, and your choice often depends on the task at hand and
personal preferences.

### Comparing Flavors

Some users have strong preferences for one flavor over another. However,
each flavor has its own merits. The choice between them depends on the
problem you're solving and your personal style.

-   Tidyverse emphasizes readability and flexibility, making it an
    excellent choice for scalable and easily comprehensible code,
    particularly when collaborating with others. It emphasizes efficient
    data cleaning and manipulation capabilities with concise and
    practical code.

-   data.table excels in speed and conciseness, allowing for rapid
    development and efficient code execution, even with sizable
    datasets.

-   Base R is known for its stability, making it a suitable choice for
    long-term projects. Additionally, it offers skills that are more
    transferable to other programming languages.

## Base R

Here is a quick look at the syntax of base R:

### R as a Calculator

R can be used as a simple calculator. You can perform basic arithmetic
operations such as addition, subtraction, multiplication, and division.

```{r}
# Example:
2 + 3
```

Variables In R, you can store values in variables. This allows you to
reuse and manipulate these values later.

```{r}
# Example:
x <- 5
y <- 3
x + y
```

### Data Frames

A data frame is a fundamental data structure in base R. It is a
two-dimensional tabular data structure with rows and columns, much like
a spreadsheet.

```{r}
# Example:
df <- data.frame(
  Name = c("Alice", "Bob", "Charlie"),
  Age = c(25, 30, 22),
  Score = c(85, 92, 78)
)
df
```

### Accessing Columns

You can access individual columns of a data frame using the `$` operator
or square brackets `[]`.

```{r}
# Example:
# Accessing the 'Name' column using the '$' operator
df$Name

# Accessing the 'Age' column using square brackets
df["Age"]
```

### Accessing Rows

You can access specific rows using indexing.

```{r}
# Example:
# Access the first row
df[1, ]

# Access the second row
df[2, ]

# Access rows 2 to 3
df[2:3, ]
```

### Conclusion

Congratulations! You've learned the basics of base R, data frames, and
how to access columns and rows within them. These skills are essential
for data analysis and manipulation in R. Keep practicing and exploring
to become more proficient in R programming.

## Tidyverse

Tidyverse, a powerful collection of R packages for data manipulation and
visualization. In this tutorial, you'll learn the basics of working with
Tidyverse, including tibbles (a modern data frame), how to view data,
summarize information, access columns, and access rows. Let's dive in!

### Tibbles

In this section, we will explore tibbles, a modern data frame structure
in the tidyverse. Tibbles are a more user-friendly and feature-rich
alternative to the traditional data frames in R.

Tibbles are a type of data frame provided by the Tidyverse. They are
similar to traditional data frames in R but offer some advantages:

-   Tibbles display data more neatly and consistently.
-   Column names are always treated as text.
-   Tibbles do not convert strings to factors by default.

Let's go over the key characteristics of tibbles:

1.  **Column Names**: Tibbles preserve column names as they are, without
    any modifications or coercion. This can be essential for maintaining
    consistency in your data.
2.  **Data Type Consistency**: Tibbles are stricter about preserving
    data types. Unlike data frames, tibbles won't automatically convert
    your variables to factors or character vectors.
3.  **Printing**: When you print a tibble to the console, it only
    displays the first few rows and columns, making it easier to explore
    large datasets.
4.  **Subsetting**: Tibbles use `[[` for subsetting columns, which can
    be more intuitive and less prone to errors.
5.  **Row Names**: Tibbles do not have row names. Instead, they use a
    row number as a default identifier.

#### Creating Tibbles

You can create a tibble from an existing data frame or by directly
specifying its columns. Let's see how to create a tibble:

```{r}
# Creating a tibble from a data frame
library(tibble)
df <- data.frame(ID = 1:3, Name = c("Alice", "Bob", "Charlie"))
tb <- as_tibble(df)

# Creating a tibble from scratch
my_tibble <- tibble(
  ID = c(101, 102, 103),
  Age = c(25, 30, 22),
  City = c("New York", "Los Angeles", "Chicago")
)
my_tibble
```

### Viewing Data

Now that we have created some tibbles, let's explore their properties
and learn how to work with them.

##### Viewing the First Few Rows

You can use the `head()` function to view the first few rows of a
tibble:

```{r}
# View the first 5 rows of my_tibble
head(my_tibble)
```

#### Viewing the Last Few Rows

Similarly, you can view the last few rows using the `tail()` function:

```{r}
# Example:
tail(my_tibble)
```

### Summary Information

#### Summary of the tibble

To get a summary of your tibble's structure, you can use `str()` or
`glimpse()`:

```{r}
# Display summary information
str(my_tibble)
glimpse(my_tibble)
```

#### Summary of data in a tibble

To get summary statistics for numerical columns, you can use the
`summary()` function:

```{r}
# Example:
summary(my_tibble$Age)
```

For categorical columns, you can use the `table()` function to see the
frequency of each category:

```{r}
# Example:
table(my_tibble$City)
```

### Accessing Columns

Accessing columns in a tibble is straightforward using the `$` operator:

```{r}
# Access the 'Age' column
my_tibble$Age
```

### Accessing Rows

Accessing data in particular rows can be done using the `filter()`
function (more on this later):

```{r}
# Filter rows where Age is greater than 25
my_tibble %>%
  filter(Age > 25)
```

Congratulations! You've learned the basics of working with Tidyverse in
R. Tibbles provide a clean and modern way to work with data, and you've
discovered how to view and summarize data, access columns, and access
rows. These skills are fundamental for data analysis and manipulation in
R, and you're on your way to becoming proficient with the Tidyverse. In
the tutorials that follow, we will dive deeper into data manipulation
using tibbles and other tidyverse tools.
